Denotational Semantics for Gradual Typing in Synthetic Guarded Domain Theory

### 2023-04-25 paper-new/intro.tex
changes to wording

% gradual typing, graduality
\subsection{Gradual Typing and Graduality}
One of the principal categories on which type systems of programming languages are classified
is that of static versus dynamic type discipline.
In static typing, the code is type-checked at compile time, while in a dynamic typing,
the type checking is deferred to run-time. Both approaches have benefits: with static typing,
the programmer is assured that if the program passes the type-checker, their program
is free of type errors. Meanwhile, dynamic typing allows the programmer to rapidly prototype
their application code without needing to commit to fixed type signatures for their functions.

\emph{Gradually-typed languages} \cite{siek-taha06} allow for both disciplines
to be used in the same codebase, and support interoperability between statically-typed
and dynamically-typed code.
This flexibility allows programmers to begin their projects in a dynamic style and
enjoy the benefits of dynamic typing related to rapid prototyping and easy modification
while the codebase ``solidifies''. Over time, as parts of the code become more mature
and the programmer is more certain of what the types should be, the code can be
\emph{gradually} migrated to a statically typed style without needing to start the project
over in a completely differnt language.

@@

\subsection{Current Approaches to Proving Graduality}
Current approaches to proving graduality include the methods of Abstracting Gradual Typing
\cite{garcia-clark-tanter2016} and the formal tools of the Gradualier \cite{cimini-siek2016}.
These allow the language developer to start with a statically typed langauge and derive a
gradually typed language that satisfies the gradual guarantee. The downside to these approaches
is that the semantics of the resulting languages are too lazy: the frameworks consider only
the $\beta$ rules and not the $\eta$ equalities. Furthermore, while these frameworks do prove
graduality, they do not show the correctness of the equational theory, which is equally important
to sound gradual typing. For example, programmers often refactor their code without thinking about
whether the refactoring has broken the semantics of the program. 
It is the validity of the laws in the equational theory that guarantees that such refactorings are sound. 
Similarly, correctness of compiler optimizations rests on the validity of the corresponding equations
from the equational theory. It is therefore important that the langages that claim to be gradually typed
have provably correct equational theories.

@@

\subsection{Proving Graduality in SGDT}
\textbf{TODO:} This section should probably be moved to after the relevant background has been introduced.

>>>>

% gradual typing, graduality
\subsection{Gradual Typing and Graduality}
One of the principal ways of categorizing type systems of programming languages is
whether they are \emph{static} or \emph{dynamic}.
In static typing, the code is type-checked at compile time, while in dynamic typing,
the type checking is deferred to run-time. Both approaches have benefits: with static typing,
the programmer is assured that if the program passes the type-checker, their program
is free of type errors. Meanwhile, dynamic typing allows the programmer to rapidly prototype
their application code without needing to commit to fixed type signatures for their functions.

\emph{Gradually-typed languages} \cite{siek-taha06} allow for both static and dynamic typing disciplines
to be used in the same codebase, and support smooth interoperability between statically-typed
and dynamically-typed code.
This flexibility allows programmers to begin their projects in a dynamic style and
enjoy the benefits of dynamic typing related to rapid prototyping and easy modification
while the codebase ``solidifies''. Over time, as parts of the code become more mature
and the programmer is more certain of what the types should be, the code can be
\emph{gradually} migrated to a statically typed style without needing to start the project
over in a completely differnt language.

@@

\subsection{Current Approaches to Proving Graduality}
Current approaches to proving graduality include the methods of Abstracting Gradual Typing
\cite{garcia-clark-tanter2016} and the formal tools of the Gradualier \cite{cimini-siek2016}.
These allow the language developer to start with a statically typed langauge and derive a
gradually typed language that satisfies the gradual guarantee. The downside to these approaches
is that the semantics of the resulting languages are too lazy: the frameworks consider only
the $\beta$ rules and not the $\eta$ equalities. Furthermore, while these frameworks do prove
graduality of the resulting languages, they do not show the correctness of the equational theory,
which is equally important to sound gradual typing. For example, programmers often refactor their
code without thinking about whether the refactoring has broken the semantics of the program. 
It is the validity of the laws in the equational theory that guarantees that such refactorings are sound. 
Similarly, correctness of compiler optimizations rests on the validity of the corresponding equations
from the equational theory. It is therefore important that the langages that claim to be gradually typed
have provably correct equational theories.

@@

\subsection{Proving Graduality in SGDT}
\textbf{TODO:} This section should probably be moved to after the relevant background has been introduced?
